Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-04-07 articles/formalism/sections/real_space_formalism.tex
Kevin revisions in section 2

%--------------------------------------------------------
%--------------------------------------------------------
%\section{The spin-0 description of CMB polarization} \label{sec:pol-intro}
%--------------------------------------------------------
\section{Polarization primer}\label{sec:pol-primer}
The CMB polarization is measured in terms of Stokes Q and U parameters. The Stokes Q is defined as the linear polarization measured along local cartesian axes ($+$ along the x-axis and $-$ along the y-axis) while Stokes U is defined as the linear polarization measured along the cartesian axes rotated by $45^{\circ}$ ($+$ along x-axis and $-$ along the y-axis in the counter-clockwise rotated system). Clearly the definitions of Stokes parameters depend on the choice of the coordinate system and it can be shown that the measurement made in two different coordinate systems rotated by an angle $\phi$ are related by the following equation,
%
\beq \label{eq:qu-rot}
\fqu' = \begin{bmatrix} \cos{2 \psi} &  \sin{2 \psi} \\ -\sin{2\psi} & \cos{2 \psi} \end{bmatrix} \fqu \,,
\eeq
%
which one can simply checked by verifying that $Q' \rightarrow U$ and $U' \rightarrow -Q$ when the coordinate systems are rotated by $45^\circ$ in the counter-clockwise sense.

Following the notation of \cite{Zaldarriaga1997}, fields which transform as,
%
\beq
{}_{s}f' = e^{-is\psi} {}_{s}f \,,
\eeq
%
under right handed rotations of the local coordinate system are termed spin-s fields. The Stokes parameters can be combined to define the complex fields,
%
\beq \label{eq:spin-pol}
_{\pm 2}\bar{X}(\hat{n}) = Q(\hat{n}) \pm i U (\hat{n}) \,, %\nonumber \\ &=& \sum_{\ell m}  {_{\pm 2}} \tilde{X}_{\ell m}  \, {_{\pm 2}}Y_{\ell m} (\hat{n}) \,.
\eeq
%
which under rotation of the local tangent plane coordinate system by an angle $\psi$ at any point on the sphere transform as: $_{\pm 2}X' = e^{\mp i2\psi} {_{\pm 2}X}$, which follows directly from \eq{eq:qu-rot}. Therefore the quantities $_{\pm2}X$ are spin${\pm2}$ fields respectively.

The Stokes parameters being coordinate dependent quantities it is inconvenient to work with them. It is possible to define spin raising ($\eth$) operator on the sphere which when operated on a field of spin-s $_{s}f$ results in a fields with spin-$(s+1)$: $(\eth _{s}f)' = e^{-i(s+1)\psi}(\eth _{s}f)$ and similarly one can define a spin lowering $(\bar{\eth})$ operators such that $(\bar{\eth} _{s}f)' = e^{-i(s-1)\psi}(\bar{\eth} _{s}f)$\cite{goldberg67}. One can now construct a complex spin-0 scalar by applying the spin raising and lowering operators on the spin-2 fields ${_{\pm 2}X}$ as follows,
%
\begin{subequations}\label{eq:ebdef}
\beqry
\mathcal{E}(\hat{n}) + i \mathcal{B}(\hat{n}) &=& -\bar{\eth}^2 _{+ 2}\bar{X}(\hat{n}) \,,\\
\mathcal{E}(\hat{n}) - i \mathcal{B}(\hat{n}) &=& -{\eth}^2 _{-2}\bar{X}(\hat{n}) \,,
\eeqry
\end{subequations}
%
which by the virtue of being scalars are independent of coordinate definitions. %The real part of this complex spin-0 scalar field corresponds to the E-mode while the imaginary part to the B-modes \cite{Kamionkowski1997}.

The complex field $_{\pm 2}X$ defined on the sphere can be decomposed in spin spherical harmonic functions: ${}_{\pm 2}X(\hat{n}) = \sum_{\ell m} {}_{\pm 2} X_{\ell m} {}_{\pm 2}Y_{\ell m}(\hat{n})$. On applying the spin raising and lowering operators on the spin spherical harmonic functions leads to the following identities\cite{goldberg67},
%
\begin{subequations}\label{eq:spinopylm} 
\beqry
\eth _s Y_{lm}(\hat{n}) &=& \sqrt{(\ell-s)(\ell+s+1)} _{s+1} Y_{lm}(\hat{n}) \,, \\
\bar{\eth} _s Y_{lm}(\hat{n}) &=& -\sqrt{(\ell+s)(\ell-s+1)} _{s-1} Y_{lm}(\hat{n}) \,, 
\eeqry
\end{subequations}
%
where $_s Y_{lm}(\hat{n}) $ denote the spin-s spherical harmonics.

Using \eq{eq:ebdef}, the spin spherical harmonic decomposition of ${}_{\pm2}X$ and the identities given in \eq{eq:spinopylm} it can be shown that the scalar fields $\mathcal{E}/\mathcal{B}$ are given by the following equations,
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) ~\,;~ \mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients $a^{E}_{\ell m}$  \& $a^{B}_{\ell m}$ are related to the harmonic coefficients of the spin-2 polarization field via the following equations,
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + {}_{-2}\tilde{X}_{\ell m} \Big] ~\,;~a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - {}_{-2}\tilde{X}_{\ell m} \Big] \,.
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields as defined by the following equations, 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n}) ~\,;~ B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
Note that these $E/B$ fields are merely filtered versions of $\mathcal{E}/\mathcal{B}$ as their spherical harmonic coefficients of expansion are related by the factor $\sqrt{\frac{(\ell+2)!}{(\ell-2)!}}$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Matrix notation} \label{sec:mat_pol_intro}
In this section we cast the relations introduced in Sec.~\ref{sec:pol-primer} in matrix notation\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}. This representation will make transparent the derivation of the real space operators we discuss in the following sections. We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.

We begin by introducing the matrices encoding the spin spherical harmonic basis vectors,
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,,
\eeq
%
where $s$ denotes the spin of the basis functions. For this work we will be working with cases $s \in [0,2]$. In this notation, each column can be mapped to a specific harmonic basis function marked by the pair of indices: $(\ell,m)$ and each row maps to a specific position on the sphere. Note that this matrix is in general not a square matrix. The number of rows is determined by the scheme used to discretely represent the sphere and the number of columns is set by the number of basis functions of interest (often determined by the band limit).

We now define the different polarization data vectors and their representation in real and harmonic space as follows,
%
\beqrys
\bar{S} &=& \bmat E \\ B  \emat_{2 \rm N_{\rm pix} \times 1} ~~~~;~~ \bar{X} = \bmat _{+2}X \\ _{-2}X \emat_{2 \rm N_{\rm pix} \times 1} ~~;~~\bar{P} =\fqu_{\tiny {2 \rm N_{\rm pix} \times 1}} \,, \\
\tilde{S} &=& \bmat a^{E} \\ a^{B} \emat _{2 \rm N_{\rm alms} \times 1}  ~~; ~~ \tilde{X} = \bmat _{+2} \tilde{X} \\ _{-2} \tilde{X} \emat_{2 \rm N_{\rm alms} \times 1} \,.
\eeqrys
%
The different symbols have the same meaning as that discussed in \sec{sec:pol-primer}, except that the subscript $_{\ell m}$ for the spherical harmonic coefficients of expansion is suppressed in favor of cleaner notation.

Next we define the operators which govern the transformations between different representations of the polarization field as follows,

@@

where we have chosen the sign conventions so as to match those used in Healpix.

@@

@@

\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.

@@

@@

Finally we introduce the harmonic space operators, which project the harmonic space data vector to E or B subspace,

@@

\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E* \tilde S \,,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B *\tilde S \,

@@

Note that these harmonic space matrices are idempotent, orthogonal to each other and their sum is an identity matrix as can be explicitly seen via the following relations, 
\begin{subequations} \label{eq:har_op_prop}
\tilde O_E * \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B * \tilde O_B= \tilde O_B \,,\\
 \tilde O_E * \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
 \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
Note that the above relations for these harmonic space operators are exactly valid.  In the following sections we aim to derive the real space analogues of these harmonic space operators.

>>>>

%--------------------------------------------------------
%--------------------------------------------------------
%\section{The spin-0 description of CMB polarization} \label{sec:pol-intro}
%--------------------------------------------------------
\section{Polarization primer}\label{sec:pol-primer}
The CMB polarization is measured in terms of Stokes $Q$ and $U$ parameters. Stokes $Q$ is defined as the linear polarization measured along local cartesian axes ($+$ along the $x$-axis and $-$ along the $y$-axis) while Stokes $U$ is defined as the linear polarization measured along the cartesian axes rotated by $45^{\circ}$ ($+$ along x-axis and $-$ along the y-axis in the counter-clockwise rotated system). The Stokes parameters depend on the choice of the coordinate system and two different coordinate systems rotated by an angle $\psi$ relate as
%
\beq \label{eq:qu-rot}
\fqu' = \begin{bmatrix} \cos{2 \psi} &  \sin{2 \psi} \\ -\sin{2\psi} & \cos{2 \psi} \end{bmatrix} \fqu \,.
\eeq
%
%which one can simply checked by verifying that $Q' \rightarrow U$ and $U' \rightarrow -Q$ when the coordinate systems are rotated by $45^\circ$ in the counter-clockwise sense.

Following the notation of \cite{Zaldarriaga1997}, fields which transform as
%
%\beq
${}_{s}f' = e^{-is\psi} {}_{s}f $
%\eeq
%
under right handed rotations of the local coordinate system are termed spin-$s$ fields. The Stokes parameters can be combined to define the complex fields,
%
\beq \label{eq:spin-pol}
_{\pm 2}\bar{X}(\hat{n}) = Q(\hat{n}) \pm i U (\hat{n}) \,, %\nonumber \\ &=& \sum_{\ell m}  {_{\pm 2}} \tilde{X}_{\ell m}  \, {_{\pm 2}}Y_{\ell m} (\hat{n}) \,.
\eeq
%
which under rotation of the local tangent plane coordinate system by an angle $\psi$ at any point on the sphere transform as: $_{\pm 2}X' = e^{\mp i2\psi} {_{\pm 2}X}$, which follows directly from \eq{eq:qu-rot}. Therefore the quantities $_{\pm2}X$ are spin ${\pm2}$ fields respectively.

The coordinate dependence of Stokes parameters make them inconvenient to work with.  With operations that raise (or lower) the spin of the field, we can produce a spin-0 scalar that is coordinate independent.   The spin-raising operator ($\eth$), when operated on a field of spin-s $_{s}f$, results in a fields with spin-$(s+1)$: $(\eth _{s}f)' = e^{-i(s+1)\psi}(\eth _{s}f)$  \cite{goldberg67}.  The complementary spin-lowering operator $(\bar{\eth})$  similarly results with $(\bar{\eth} _{s}f)' = e^{-i(s-1)\psi}(\bar{\eth} _{s}f)$.  The complex spin-0 scalar now arise from  the spin-2 fields ${_{\pm 2}X}$ as follows,
%
\begin{subequations}\label{eq:ebdef}
\beqry
\mathcal{E}(\hat{n}) + i \mathcal{B}(\hat{n}) &=& -\bar{\eth}^2 _{+ 2}\bar{X}(\hat{n}) \,,\\
\mathcal{E}(\hat{n}) - i \mathcal{B}(\hat{n}) &=& -{\eth}^2 _{-2}\bar{X}(\hat{n}) \,.
\eeqry
\end{subequations}
%
%which by the virtue of being scalars are independent of coordinate definitions. %The real part of this complex spin-0 scalar field corresponds to the E-mode while the imaginary part to the B-modes \cite{Kamionkowski1997}.

The complex field $_{\pm 2}X$ defined on the sphere can be decomposed in spin spherical harmonic functions: ${}_{\pm 2}X(\hat{n}) = \sum_{\ell m} {}_{\pm 2} X_{\ell m} {}_{\pm 2}Y_{\ell m}(\hat{n})$. On applying the spin raising and lowering operators on the spin spherical harmonic functions leads to the following identities \cite{goldberg67},
%
\begin{subequations}\label{eq:spinopylm} 
\beqry
\eth _s Y_{lm}(\hat{n}) &=& \sqrt{(\ell-s)(\ell+s+1)} _{s+1} Y_{lm}(\hat{n}) \,, \\
\bar{\eth} _s Y_{lm}(\hat{n}) &=& -\sqrt{(\ell+s)(\ell-s+1)} _{s-1} Y_{lm}(\hat{n}) \,, 
\eeqry
\end{subequations}
%
where $_s Y_{lm}(\hat{n}) $ denote the spin-s spherical harmonics.

Using the definition of $\mathcal{E/B}$, the spin spherical harmonic decomposition of ${}_{\pm2}X$ and the identities given in \eq{eq:spinopylm} it can be shown that the scalar fields $\mathcal{E}/\mathcal{B}$ are given by the following equations,
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) ~\,;~ \mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients $a^{E}_{\ell m}$  \& $a^{B}_{\ell m}$ are related to the harmonic coefficients of the spin-2 polarization field via the following equations,
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + {}_{-2}\tilde{X}_{\ell m} \Big] ~\,;~a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - {}_{-2}\tilde{X}_{\ell m} \Big] \,.
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields as defined by the following equations, 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n}) ~\,;~ B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
Note that these $E/B$ fields are merely filtered versions of $\mathcal{E}/\mathcal{B}$ as their spherical harmonic coefficients of expansion are related by the factor $[{(\ell+2)!}/{(\ell-2)!}]^{1/2}$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Matrix notation} \label{sec:mat_pol_intro}

Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors,
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,%,
\eeq
%
%where $s$ denotes the spin of the basis functions.
We will be working with cases $s \in [0,2]$. Each column maps to a specific harmonic basis function (i.e. indexed by $\ell m$) and each row maps to a pixel on the sphere. This matrix is not square in general: the number of rows is determined by the pixelization and the number of columns is set by the number of basis functions (e.g. determined by the band limit).

We now define the different polarization data vectors and their representation in real  space as and harmonic as follows,\footnote{We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.}
%
\beqrys
\bar{S} &=& \bmat E \\ B  \emat_{2 \rm N_{\rm pix} \times 1} ~~~~;~~ \bar{X} = \bmat _{+2}X \\ _{-2}X \emat_{2 \rm N_{\rm pix} \times 1} ~~;~~\bar{P} =\fqu_{\tiny {2 \rm N_{\rm pix} \times 1}} \,, \\
\tilde{S} &=& \bmat a^{E} \\ a^{B} \emat _{2 \rm N_{\rm alms} \times 1}  ~~; ~~ \tilde{X} = \bmat _{+2} \tilde{X} \\ _{-2} \tilde{X} \emat_{2 \rm N_{\rm alms} \times 1} \,.
\eeqrys
%
The different symbols have the same meaning as that discussed in \sec{sec:pol-primer}, except that the subscript ${\ell m}$ for the spherical harmonic coefficients is suppressed for cleaner notation.

We define transformations between different representations of the polarization field (i.e. from $Q,U$ to $_{\pm2} X$):

@@

The sign conventions we have chosen match HEALPix.

@@

  %Kevin's version

@@

  \bar{X} &= \bar T  \bar{P}; &\qquad \bar{P} = \frac{1}{2} \bar T^{\dagger}  \bar{X} ; \\
  \tilde{X} &= \tilde T \tilde{S}; &\qquad \tilde{S} = \frac{1}{2}\tilde T^{\dagger} \tilde{X} \,.
  Meanwhile the spherical harmonic transforms are written as:
  \bar X &=  {{}_2\mathcal{Y}}  \tilde X; &\qquad \tilde X ={{}_2\mathcal{Y}}^{\dagger}  \bar X  ; \\
  \bar S &=  {{}_0\mathcal{Y}} \tilde S; &\qquad  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} \bar S \,.
% Aditya's version
%\beqry
%\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
%\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
%\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
%\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.

@@

%\eeqry
Finally we introduce the operators that project harmonic space data vector to the $E$ or $B$ subspace,
\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E  \tilde S ,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B  \tilde S .

@@

Note that these harmonic space matrices are idempotent ($\tilde O_E  \tilde O_E = \tilde O_E;  \tilde O_B  \tilde O_B= \tilde O_B$), orthogonal ($\tilde O_E  \tilde O_B = \mathbb{0}$), and sum to the identity matrix ($\tilde O_E + \tilde O_B = \mathbb{1}$).

@@

%\begin{subequations} \label{eq:har_op_prop}
%\beqry
%\tilde O_E  \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B  \tilde O_B= \tilde O_B \,,\\
% \tilde O_E  \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
% \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
%\eeqry
%\end{subequations}

@@

The above relations for these harmonic space operators are exactly valid.  In the following sections we aim to derive the real space analogues ($O_E,O_B$) of these harmonic space operators.
